Limited evidence for widespread heterogeneity in psychology

### 2018-07-10 writing/draft-heterogeneity.rmd
Updated draft and associated files

affiliation:
  - id            : "1"
    institution   : "Tilburg University"
  - id            : "2"
    institution   : "Utrecht University"

abstract: |
  We examined the evidence for heterogeneity (of effect sizes) and explored the correlation between heterogeneity and effect size in a sample of ten pre-registered multi-lab replication projects in psychology (Many Labs 1 & 3, Registered Replication Reports 1 to 8). Across the 37 main outcomes of these projects we found limited heterogeneity, only 4/37 (11%) effects had significant heterogeniety and the majority (32/37; 86%) were most likely to have zero to small heterogeneity. Power to detect small heterogeneity, as defined by @10.1136/bmj.327.7414.557, was for all projects very low (mean 36%), but good to excellent for medium and large heterogeneity. We saw a strong correlation between observed effect size and heterogeneity in our sample (R^2^ = .61). Our results suggest that an association between heterogeneity and effect size may need to be considered when estimating heterogeneity, and a lack of evidence to indicate widespread heterogeneity in psychology.

keywords          : "heterogeneity, meta-analysis, replication, psychology, many labs, registered replication reports"
wordcount         : "146"

@@

source("../code/helper_functions_tables_figures.r") #Load functions to prep data for figures
```
Empirical research is typically portrayed as proceeding in two stages. First, belief in the existence of an effect is established. Second, the effect's generalizability is examined by exploring its boundary conditions. In the first stage, inferential statistics are used  to minimize the risk that a discovery is due to sampling error.  In the second stage, one may ask to what extent the effect depends on a particular choice of four contextual factors; the 1) sample population, 2) settings, 3) treatment variables and 4) measurement variables [@Campbell2015]. This extent is often explored through replications of the original study that are either as similar as possible to the original (called 'direct' or 'exact' replications) or with some deliberate variation on conceptual factors [so-called 'conceptual' or 'indirect' replications; @10.1017/S0140525X17001972], and once sufficient studies have accumulated through meta-analysis. In meta-analysis, the heterogeneity of an effect size (henceforth referred to as heterogeneity) is a measure of an effect's susceptibility to changes in these four factors. An effect strongly dependent on one or more of the four factors, unless controlled for, should exhibit high heterogeneity. In this paper we examine the heterogeneity in replication studies in psychology , focusing on direct replications, and explore a proposed relationship between effect size and heterogeneity.

Heterogeneity is of concern for several reasons. First and foremost, unaccounted for heterogeneity can have practical consequences not to be ignored. This is readily evident for medicine, where in the case of heterogeneity an intervention, such as a medication, that is successful for some may have direct negative health consequences for others. The same is true of mental health interventions in psychology, but heterogeneity can also have major consequences for topics such as child development, education, and business performance, where psychological research often impacts policy recommendations. Thus, heterogeneity should be no less of a concern for psychologists than for medical practitioners.

Second, unaccounted for heterogeneity is an indication of incomplete theory, since it suggests that a theory is unable to predict all contextual factors of importance to its claims. As such, heterogeneity might imply previously unknown predictors, so called 'hidden moderators' [@10.2139/ssrn.2820883], the discovery of which can be seen as an opportunity for theoretical advancement [@Tackett2017].

Third,  the possibility of heterogeneity can create controversy in the interpretation of replication results. The proclamation of a  'failure' to replicate an effect is sometimes taken to imply that the original find was merely a statistical fluke, due to 'p-hacking' [@10.1177/0956797611417632] or publication bias [@10.2139/ssrn.2659409]. Unsurprisingly, some researchers take offense [e.g., @10.1016/j.jesp.2016.02.003]. An alternative explanation for non-replication, often espoused by the original authors [e.g, @10.2139/ssrn.2698810; @10.1177/1745691616674460], is that the effect is more heterogeneous than (perhaps implicitly) claimed originally. Such explanations may be valid or not, but even if valid, an effect is typically of less general interest the more specific circumstances it requires to appear. To attenuate the risk of heated discussions on the (non)existence of an effect, original authors may wish to consider pre-specifying the degree of heterogeneity that would make even them lose interest in the effect (e.g., by declaring 'constraints on generality' [@10.1177/1745691617708630]. To conclude, heterogeneity or its absence provides vital information for the implementation of research in practice, the advancement of theory, and the interpretation of research outcomes.

Heterogeneity also affects meta-analytic techniques used to statistically summarize findings on a certain topic. Techniques that attempt to correct for publication bias in their estimate tend to fail in the presence of heterogeneity [@10.1177/1745691616662243; @10.1037/met0000025; @10.1177/1745691616650874], which is problematic considering widespread publication bias [@10.1007/s11192-011-0494-7] and its impact on meta-analytic estimates. In addition, heterogeneity alters the interpretation of meta-analytic estimates as either _the_ true effect size (under homogeneity) or the average of the true effect sizes (under heterogeneity), though one may question the usefulness of interpreting the average true effect size in the presence of heterogeneity [@Simonsohn2017], just as it may be questionable to interpret an average main effect in the context of an interaction effect [@Aiken1991].

It is a commonly believed that heterogeneity is the norm in psychology.  In support of this notion, recent large scale reviews of meta-analyses in psychology [@Stanley2017; @10.5334/jopd.33] report median heterogeneity levels that can best be described as 'large' [see next paragraph; @10.1136/bmj.327.7414.557]. In comparison, the median heterogeneity estimate in medicine [@10.1136/bmj.39343.408449.80] would be considered 'small' by the same standard. It may simply be that effects in psychology are more heterogeneous than those of medicine. However, meta-analyses in psychology also typically include more studies than those in medicine, and it could be that they tend to include studies from a much broader spectrum. The median number of studies (effects) per meta-analysis in the psychology sample of @10.5334/jopd.33 was 12, whereas in medicine it is only 3 [@10.1186/1471-2288-11-160]. It is difficult to separate these explanations (intrinsically more heterogeneity, or psychology simply including studies from a broader spectrum?). To facilitate doing so, in this paper we focus on meta-analyses of only direct replications, which are exempt from the potential problem of including too disparate studies. In reference to meta-analyses of direct replications, several authors [@10.1177/1745691616662243; @Tackett2017] have argued that if we were to expect heterogeneity to be absent or minimal anywhere, it would be in pre-registered multi-lab projects with a common protocol [such as @10.1027/1864-9335/a000178]. They further argue that the fact that heterogeneity has been reported even under such circumstances is an indication of widespread heterogeneity in psychology. However, even in the case of multi-lab direct replication projects, studies will still vary on two contextual factors (sample population and settings) and if we believe an effect is sensitive to changes in these two factors we might also expect to find some heterogeneity.

@@

One problem lies in the assessment of heterogeneity and its inherent uncertainty. Heterogeneity is often measured by the $I^2$ index [@10.1002/sim.1186;  @10.1136/bmj.327.7414.557], which allows comparison of estimates across meta-analyses and has an intuitive interpretation. It can be interpreted as the percentage of variability in observed effect sizes in a meta-analysis that is due to heterogeneity amongst the true effects (that is, sensitivity to contextual factors) rather than sampling variance, and ranges from 0-100%. The $I^2$ index is a transformation of the well-known Q-test of heterogeneity in meta-analysis [@10.2307/3001666] by taking $[Q - (k - 1)] / Q \times 100$%, where k is the number of studies included in the meta-analysis and a negative $I^2$ is set to zero. @10.1136/bmj.327.7414.557 tentatively defined $I^2$-values of 25, 50, and 75% as small/medium/large heterogeneity respectively. Tests of heterogeneity typically have low statistical power in many practical situations [@10.1002/sim.2481; @10.1037/1082-989X.11.2.193]. This complicates the discussion of heterogeneity, because while $I^2$ always provides an estimate of heterogeneity, this estimate is often accompanied by wide confidence intervals [@10.1136/bmj.39343.408449.80]. For example, Ioannidis reports that in a large set of Cochrane meta-analyses, all meta-analyses with $I^2$ point estimates of 0% had upper 95% confidence intervals that exceeded 33%, exceeding what @10.1136/bmj.327.7414.557 defined as 'small' heterogeneity. In addition, under homogeneity $I^2$ has a central chisquare distribution [@10.1186/s12874-015-0024-z], a distribution that is right-skewed with more than 40% of observations falling above the expected value (for all k > 4). In other words, even in the absence of heterogeneity, a meta-analysis of 5 or more studies will have an $I^2$ point estimate above zero in more than 40% of cases. Heterogeneity estimates may thus be congruent with a wide range of true heterogeneity levels. Despite exhortations to the contrary [@10.1136/bmj.39343.408449.80], it remains common to omit confidence intervals in the reporting of $I^2$. In consideration of such uncertainty and the prevalent belief that heterogeneity is the norm in psychology, we examine the existing evidence for heterogeneity in psychology using a sample of pre-registered multi-lab direct replication projects.

Effect size is likely associated with heterogeneity. Intuitively, it makes sense to believe that if there is no meta-analytic effect there is nothing to moderate (i.e., no heterogeneity). However, a null or near null effect size estimate may arise from failure to consider contextual factors ('hidden moderators'; @10.2139/ssrn.2820883) and does not by itself imply the absence of heterogeneity. A large meta-analytic effect size on the other hand, can be expected to be associated with more heterogeneity. To illustrate, consider a meta-analysis of say, the correlation between neuroticism and procrastination. Each included study would need to measure the two variables somehow, possibly the same way across studies. However, because of individual differences and differences in study samples, measurement reliabilities may differ across studies either due to sampling variance (that the sample happens to be more or less homogeneous) or to differences in contextual factors (e.g., sampling population, method of measurement). This means that even if the underlying true effect size is the same, the observed correlation between the two variables will differ between studies [see also @Schmidt2015]. Keeping measurement reliabilities constant, differences in observed effect sizes will increase with the underlying true effect, resulting in more variability being ascribed to heterogeneity. More formally, an observed correlation $r_{xy}$ can be expressed as the product of the true correlation or effect size, $\rho_{xy}$, multiplied by the square root of the measurement reliabilities for X ($R_{xx'}$) and Y ($R_{yy'}$): $r_{xy} = \rho_{xy} \times \sqrt{R_{xx'}} \times \sqrt{R_{yy'}}$. As such, keeping constant study differences in $\sqrt{R_{xx'}} \times \sqrt{R_{yy'}}$ while increasing true effect size $\rho_{xy}$ increases the observed differences between studies, thereby increasing heterogeneity of observed effect sizes (see Table 1). We explore the association between effect size and heterogeneity in a sample of pre-registered multi-lab replication projects in psychology.

@@

het_demo %>% knitr::kable(col.names = c("", "Rho = 0", "Rho = .3", "Rho = .5"), digits = 2)

@@

We examine the evidence for widespread heterogeneity in psychology and explore the correlation between effect size and heterogeneity, using ten pre-registered multi-lab replication projects in psychology (Table 2). These ten projects examined a total of 37 primary outcome variables and arguably represent the best, least biased, meta-analytic data currently available in psychology. To better interpret the heterogeneity estimates we also estimate power of each project to find zero/small/medium/large heterogeneity as defined by @10.1136/bmj.327.7414.557. Consequently, our analyses will provide information on how two contextual factors (sample population and settings) may affect consistency or heterogeneity of effects in psychology.

@@

All code and data for this project are available on the Open Science Framework (OSF) at osf.io/4z3e7/. We refer to relevant files on OSF using brackets and links in the sections below, that is (osf.io/XXXX/). We ran all analyses using R version 3.4.3 [@RCoreTeam2017].

We downloaded and collated summary data from ten pre-registered multi-lab replication projects in psychology (Table 2). Data from all ten projects were publicly available on the Open Science Framework (osf.io) and downloaded between 2018/02/01 and 2018/03/31. Although some projects (e.g. RRR4) reported results from several outcome variables, we only included primary outcome variables as stated in accompanying publications, resulting in a total of 37 effects. For each effect we extracted (osf.io/3bmvc/) summary data (e.g., means and standard deviations) at the level of the lab as specified by the original authors for their primary analysis (i.e., typically after exclusions). We extracted information on the country of each lab, whether participants were physically in the lab for the study, total number of participants per lab, type of effect size, and additional information related to each effect (see codebook; osf.io/uhq4r/). Extracted data were in a variety of formats: Excel (Many labs 1, RRR1 & RRR2), CSV (Many labs 3, RRR3, RRR4, RRR5, RRR6) and as PDF tables (RRR7, RRR8). In two cases (RRR5 and RRR6) it was necessary to download the raw data to extract summary data. Although a particular lab may have participated in several projects, the lab indicator was typically not the same across projects. Even so, we kept the original lab indicators to facilitate comparing observations in our dataset with the original datasets. Finally, we collated the summary data for all effects into one dataset for analysis (osf.io/456fz/).

To examine heterogeneity of each of the 37 effects, we computed meta-analytic estimates for all 37 effects in our dataset (Table 3). We ran all analyses as specified by the replication authors (osf.io/kf6pt/). All effects were estimated with random-effects models and the Restricted Maximum Likelihood (REML) estimator using the R-package metafor [@10.18637/jss.v036.i03], though with a variety of outcome variables: correlations, standardized mean differences, raw mean differences, odds ratios and risk differences. ML1 transformed odd ratios into standardized mean differences when meta-analyzing under the assumption that responses followed logistic distributions [@10.1037/1082-989X.8.4.448; @10.18637/jss.v036.i03]. Two projects (RRR5 and RRR7) used the Knapp and Hartung adjustment of the standard errors [@10.1002/sim.1482] and ML3 correlations were corrected for bias [@10.1037//0021-9010.74.3.469; @10.18637/jss.v036.i03]. After estimating effects, ML3 transformed correlations into eta-squared for reporting, which we did not. For each effect we estimated $I^2$ and its 95% confidence interval. Confidence intervals were calculated in metafor which uses the Q-profile method [@10.1186/1471-2288-14-103].

In order to facilitate interpretation of our results, we estimated type I error and power of the Q-test of heterogeneity for each of the 37 effects under zero/small/medium/large heterogeneity ($I^2$ = 0/25/50/75%  respectively). In addition, we plotted $I^2$-distributions across effects at these four heterogeneity levels and compared them with the distribution of the observed $I^2$ estimates of the 37 effects. To do so we simulated results for each effect given the number of studies, k, and the sample sizes, $N_k$, of those studies (osf.io/gbf4u/). We simulated the distribution of $I^2$ for correlation, standardized mean difference, and mean difference effect size measures. Because the standard error of the observed effect size in case of odds ratios and risk differences is strongly affected by assumptions of the probability of success in both treatment groups, we did not directly simulate the $I^2$ distribution for these effect size measures. In these cases, we used the information on study sample sizes and treated the effects as mean differences in our simulations.

As our concern was heterogeneity, for convenience we set the true effect size to zero in our simulations. This should not affect the results for correlations or mean differences, as estimates of effect size and heterogeneity for these measures are statistically independent. For standardized mean differences we expect negligible effects on the results, because while these estimates of effect size are positively correlated to their standard errors, the within study variance $\sigma^2$ was kept constant across studies. As a sensitivity test we ran the analysis with 'medium' effect sizes [@Cohen1988] and found the same average power at the different heterogeneity levels, see Appendix A (osf.io/bsjhu/).

In case the observed effect was a correlation, one run of a simulation proceeded as follows. First, we randomly sampled K Fisher-transformed true correlations $\rho_k$ from a normal distribution with mean 0 and heterogeneity (variance) $\tau^2$. Fisher's z-transformation is a normalizing correlation transformation that ranges from negative infinity to positive infinity, though except for extreme correlations it stays close to the -1 to 1 range [@10.2307/2331838; @Fisher1921]. Second, for each of the K true Fisher-transformed correlations we sampled one Fisher-transformed observed correlation from a normal distribution with mean $\rho_k$ and variance $1 /(N_k - 3)$. Finally, we fitted a random-effects meta-analysis with REML and estimated $I^2$ for that run. In the simulations, we varied the between-studies standard deviation $\tau$ between 0.000 and 0.50 in increments of 0.005, and used 1,000 runs at each step to approximate the distribution of $I^2$ at that value for true heterogeneity.

For mean differences we assumed a within-study variance of one for both treatment and control groups, $\sigma^2_c = \sigma^2_t = 1$. For each run we then set the population mean of the control condition to 0 and sampled K treatment population means $\mu_k$ from $N(0, \tau)$. Subsequently, K sample means for both control and treatment conditions were sampled, with $\bar{x}_c \sim N(0, 1 / \sqrt{n_c})$ and $\bar{x}_t \sim N(\mu_k, 1 / \sqrt{n_t})$, where $n_t$ and $n_c$ were the observed treatment and control sample sizes for each study. Group variances were sampled using $s^2_c \sim \chi^2 (n_c - 1) / (n_c - 1)$ and  $s^2_t \sim \chi^2 (n_t - 1) / (n_t - 1)$. Finally, we fitted a random-effects meta-analysis with REML and estimated $I^2$ for that run. For standardized mean differences we proceeded identically, except that we in the final step asked metafor to transform the effect size into a standardized mean difference in fitting the random-effects model. As with correlations, the distribution of $I^2$ was approximated for values of $\tau$ from 0 to .5 in steps of .005, using 1,000 runs at each step.

To approximate the statistical power of all 37 effects at zero, small, medium, and large heterogeneity we continued as follows. For each of the 37 effects we selected the values of $\tau$ which yielded the average value of $I^2$ in the simulations closest to 25 (small), 50 (medium), and 75 (large). For these values of $\tau$ and for $\tau = 0$ (homogeneity) we again ran 10,000 simulations, and for each run $I^2$ was calculated and the Q-test of heterogeneity was performed, yielding estimates of type I error (in case of homogeneity) and power (for heterogeneity) for each of the 37 effects. We considered a result significant when _p_ $\leq 0.05$ for the Q-test. The distributions of $I^2$ for zero, small, medium, large heterogeneity, which we compared to the observed distribution of 37 effect sizes, was generated by pooling the 37 distributions of 10,000 $I^2$ values in each category of heterogeneity. Hence these $I^2$ distributions can be considered a mixture distribution of 37 distributions, using equal weights across all 37 effects.

>>>>

affiliation:
  - id            : "1"
    institution   : "Department of Methodology and Statistics, Tilburg School of Social and Behavioral Sciences, Tilburg University, the Netherlands"
  - id            : "2"
    institution   : "Department of Sociology, Faculty of Social and Behavioural Sciences, Utrecht University, the Netherlands"

abstract: |
  We examined the evidence for heterogeneity (of effect sizes) and explored the correlation between heterogeneity and effect size in a sample of 37 effect sizes from ten pre-registered multi-lab replication projects in psychology. We found limited heterogeneity; only 4/37 (11%) effects had significant heterogeneity, and most effects (32/37; 86%) were most likely to have zero to small heterogeneity. Power to detect small heterogeneity, was low for all projects (mean 36%), but good to excellent for medium and large heterogeneity. As anticipated, we found a strong correlation between observed effect size and heterogeneity in our sample (R^2^ = .61). Our results suggest that an association between heterogeneity and effect size may need to be considered when estimating heterogeneity, and a lack of evidence to indicate widespread heterogeneity in psychology.

keywords          : "heterogeneity, meta-analysis, replication, psychology, many labs, registered replication reports"
wordcount         : "128"

@@

source("../code/helper_functions_tables_figures.r") #Load functions to prep data for figures
```
Empirical research is typically portrayed as proceeding in two stages. First, belief in the existence of an effect is established. Second, the effect's generalizability is examined by exploring its boundary conditions [@10.1177/1745691617708630]. In the first stage, inferential statistics are used  to minimize the risk that a discovery is due to sampling error.  In the second stage, one may ask to what extent the effect depends on a particular choice of four contextual factors; the 1) sample population, 2) settings, 3) treatment variables and 4) measurement variables [e.g., @Campbell2015]. This extent is often explored through replications of the original study that are either as similar as possible to the original (called 'direct' or 'exact' replications) or with some deliberate variation on conceptual factors [so-called 'conceptual' or 'indirect' replications; @10.1017/S0140525X17001972], and once sufficient studies have accumulated through meta-analysis. In meta-analysis, the heterogeneity of an effect size (henceforth referred to as heterogeneity) is a measure of an effect's susceptibility to changes in these four factors. An effect strongly dependent on one or more of the four factors, unless controlled for, should exhibit high heterogeneity. In this paper we examine the heterogeneity in replication studies in psychology , focusing on direct replications, and explore a proposed relationship between effect size and heterogeneity.

Heterogeneity is of concern for several reasons. First and foremost, unaccounted for heterogeneity can have practical consequences not to be ignored. This is readily evident for medicine, where in the case of heterogeneity an intervention, such as a medication, that is successful for some may have direct negative health consequences for others. The same is true of mental health interventions in psychology, but heterogeneity can also have major consequences for topics such as child development, education, and business performance, where research often impacts policy recommendations. Thus, heterogeneity should be no less of a concern for psychologists than for medical practitioners.

Second, unaccounted for heterogeneity is an indication of incomplete theory, since it suggests that a theory is unable to predict all contextual factors of importance to its claims. As such, heterogeneity might imply previously unknown predictors, so called 'hidden moderators' [@10.2139/ssrn.2820883], the discovery of which can be seen as an opportunity for theoretical advancement [@Tackett2017; @10.1177/1745691617708630].

Third, the possibility of heterogeneity can create controversy in the interpretation of replication results. The proclamation of a 'failure' to replicate an effect is sometimes taken to suggest that the original finding was merely a statistical fluke, due to 'p-hacking' [@10.1177/0956797611417632] or publication bias [@10.2139/ssrn.2659409]. Unsurprisingly, some researchers take offense [e.g., @10.1016/j.jesp.2016.02.003], interpreting such implications as attacks on their abilities as researchers. An alternative explanation for non-replication, often espoused by the original authors [e.g, @10.2139/ssrn.2698810; @10.1177/1745691616674460], is that the effect is more heterogeneous than (perhaps implicitly) claimed originally. Such explanations may be valid or not, but even if valid, an effect is typically of less general interest the more specific circumstances it requires to appear. To attenuate the risk of heated discussions on the (non)existence of an effect, original authors are recommended to consider pre-specifying the degree of heterogeneity that would make even them lose interest in the effect (e.g., by declaring 'constraints on generality' [@10.1177/1745691617708630]. To conclude, heterogeneity or its absence provides vital information for the implementation of research in practice, the advancement of theory, and the interpretation of research outcomes.

Heterogeneity also affects meta-analytic techniques used to statistically summarize findings on a certain topic. Techniques that attempt to correct for publication bias in their estimate tend to fail in the presence of heterogeneity [@10.1177/1745691616662243; @10.1037/met0000025; @10.1177/1745691616650874], which is problematic considering the supposedly widespread publication bias [@10.1007/s11192-011-0494-7] and its impact on meta-analytic estimates. In addition, heterogeneity alters the interpretation of meta-analytic estimates as either _the_ true effect size (under homogeneity) or the average of the true effect sizes (under heterogeneity), though one may question the usefulness of interpreting the average true effect size in the presence of heterogeneity [@Simonsohn2017], just as it may be questionable to interpret an average main effect in the context of an interaction effect [@Aiken1991].

It is a commonly believed that heterogeneity is the norm in psychology.  In support of this notion, recent large scale reviews of meta-analyses in psychology [@Stanley2017; @10.5334/jopd.33] report median heterogeneity levels that can best be described as 'large' [see next paragraph; @10.1136/bmj.327.7414.557]. In comparison, the median heterogeneity estimate in medicine [@10.1136/bmj.39343.408449.80] would be considered 'small' by the same standard. It may simply be that effects in psychology are more heterogeneous than those of medicine. However, meta-analyses in psychology also typically include more studies than those in medicine, and it could be that they tend to include studies from a much broader spectrum, that is, varying on more contextual factors (sample population, settings, treatment variables, measurement variables) or varying more on these four factors. The median number of studies (effects) per meta-analysis in the psychology sample of @10.5334/jopd.33 was 12, whereas in medicine it is only 3 [@10.1186/1471-2288-11-160]. It is difficult to separate these explanations (intrinsically more heterogeneity, or psychology including studies from a broader spectrum?). To facilitate doing so, in this paper we focus on meta-analyses of only direct replications, which are exempt from the potential problem of including too disparate studies. In reference to meta-analyses of direct replications, several authors [@10.1177/1745691616662243; @Tackett2017] have argued that if we were to expect heterogeneity to be absent or minimal anywhere, it would be in pre-registered multi-lab projects with a common protocol [such as @10.1027/1864-9335/a000178]. They further argue that the fact that heterogeneity has been reported even under such circumstances is an indication of widespread heterogeneity in psychology. However, even in the case of multi-lab direct replication projects, studies will still vary on two contextual factors (sample population and settings) and if we believe an effect is sensitive to changes in these two factors we might also expect to find some heterogeneity.

@@

One problem lies in the assessment of heterogeneity and its inherent uncertainty. Heterogeneity is often measured by the _I^2^_ index [@10.1002/sim.1186;  @10.1136/bmj.327.7414.557], which allows comparison of estimates across meta-analyses and has an intuitive interpretation. It can be interpreted as the percentage of variability in observed effect sizes in a meta-analysis that is due to heterogeneity amongst the true effects (that is, sensitivity to contextual factors) rather than sampling variance, and ranges from 0-100%. The _I^2^_ index is a transformation of the well-known Q-test of heterogeneity in meta-analysis [@10.2307/3001666] by taking $[Q - (k - 1)] / Q \times 100$%, where k is the number of studies included in the meta-analysis and a negative _I^2^_ is set to zero. @10.1136/bmj.327.7414.557 tentatively defined _I^2^_-values of 25, 50, and 75% as small/medium/large heterogeneity respectively, labels we also use un this paper.

Tests of heterogeneity typically have low statistical power in many practical situations [@10.1002/sim.2481; @10.1037/1082-989X.11.2.193]. This complicates the discussion of heterogeneity, because while _I^2^_ always provides an estimate of heterogeneity, this estimate is often accompanied by wide confidence intervals [@10.1136/bmj.39343.408449.80]. For example, Ioannidis reports that in a large set of Cochrane meta-analyses, all meta-analyses with _I^2^_ point estimates of 0% had upper 95% confidence intervals that exceeded _I^2^_ estimates of 33%, exceeding what @10.1136/bmj.327.7414.557 defined as 'small' heterogeneity. In addition, under homogeneity _I^2^_ has a central chisquare distribution [@10.1186/s12874-015-0024-z], a distribution that is right-skewed with more than 40% of observations falling above the expected value (for all k > 4). In other words, even in the absence of true heterogeneity, a meta-analysis of 5 or more studies will have an _I^2^_ point estimate above zero in more than 40% of cases. Heterogeneity estimates may thus be congruent with a wide range of true heterogeneity levels. Despite exhortations to the contrary [@10.1136/bmj.39343.408449.80], it remains common to omit confidence intervals in the reporting of _I^2^_. In consideration of such uncertainty and the prevalent belief that heterogeneity is the norm in psychology, we examine the existing evidence for heterogeneity in psychology using a sample of pre-registered multi-lab direct replication projects.

###Association between effect size and heterogeneity

Effect size is likely associated with heterogeneity. Intuitively, it makes sense to believe that if there is no meta-analytic effect there is nothing to moderate (i.e., no heterogeneity). However, a null or near null effect size estimate may arise from failure to consider contextual factors ('hidden moderators'; @10.2139/ssrn.2820883) and does not by itself imply the absence of heterogeneity. A large meta-analytic effect size on the other hand, can be expected to be associated with more heterogeneity. To illustrate, consider a meta-analysis of say, the correlation between neuroticism and procrastination. Each included study would need to measure the two variables somehow, possibly the same way across studies in the meta-analysis. However, because of individual differences and differences in study samples, measurement reliabilities may differ across studies either due to sampling variance (that the sample happens to be more or less homogeneous) or to differences in contextual factors (e.g., sampling population, measurement variables). This means that even if the underlying true effect size is the same, the observed correlation between the two variables will differ between studies [see also @Schmidt2015]. Keeping measurement reliabilities constant, differences in observed effect sizes will increase with the underlying true effect, resulting in more variability being ascribed to heterogeneity. More formally, an observed correlation $r_{xy}$ can be expressed as the product of the true correlation or effect size, $\rho_{xy}$, multiplied by the square root of the measurement reliabilities for X ($R_{xx'}$) and Y ($R_{yy'}$): $r_{xy} = \rho_{xy} \times \sqrt{R_{xx'}} \times \sqrt{R_{yy'}}$. As such, keeping constant study differences in $\sqrt{R_{xx'}} \times \sqrt{R_{yy'}}$ while increasing true effect size $\rho_{xy}$ increases the observed differences between studies, thereby increasing heterogeneity of observed effect sizes (see Table 1). We therefore explore with a meta-meta-analysis if a positive association exists between effect size and heterogeneity in the sample of pre-registered multi-lab replication projects in psychology.

@@

het_demo %>% knitr::kable(col.names = c("Effect size", "Rho = 0", "Rho = .3", "Rho = .5"), digits = 2)

@@

Table 2 lists the ten replication projects, with a total of 37 primary outcome variables, we used to examine heterogeneity and the correlation between effect size and heterogeneity in psychology. As all ten projects were (relatively) large-scale and pre-registered, our dataset arguably represents the best, least biased, meta-analytic data currently available in psychology. To better interpret the heterogeneity estimates we also estimate power of each project to find zero/small/medium/large heterogeneity. Consequently, our analyses will provide information on how two contextual factors (sample population and settings) may affect consistency or heterogeneity of effects in psychology, and on the precision of its estimate.

@@

All code and data for this project are available on the Open Science Framework (OSF) at osf.io/4z3e7/. We refer directly to relevant files on the OSF using brackets and links in the sections below. We ran all analyses using R version 3.4.3 [@RCoreTeam2017].

We downloaded and collated summary data from the ten pre-registered multi-lab replication projects in psychology (Table 2). Data from all ten projects were publicly available on the Open Science Framework (osf.io) and downloaded between 2018/02/01 and 2018/03/31. Although some projects (e.g. RRR4) reported results from several outcome variables, we only included primary outcome variables as stated in accompanying publications, resulting in a total of 37 effects. For each effect we extracted (osf.io/3bmvc/) summary data (e.g., means and standard deviations) at the level of the lab as specified by the original authors for their primary analysis (i.e., typically after exclusions). We extracted information on the country of each lab, whether participants were physically in the lab for the study, total number of participants per lab, type of effect size, and additional information related to each effect (see codebook; osf.io/uhq4r/). Extracted data were in a variety of formats: Excel (Many labs 1, RRR1 & RRR2), CSV (Many labs 3, RRR3, RRR4, RRR5, RRR6) and as PDF tables (RRR7, RRR8). In two cases (RRR5 and RRR6) it was necessary to download the raw data to extract summary data. Although a particular lab may have participated in several projects, the lab indicator was typically not the same across projects. Even so, we kept the original lab indicators to facilitate comparing observations in our dataset with the original datasets. Finally, we collated the summary data for all effects into one dataset for analysis (osf.io/456fz/).

To examine heterogeneity of each of the 37 effects, we computed meta-analytic estimates for all 37 effects in our dataset (Table 3). We ran all analyses as specified by the replication authors (osf.io/kf6pt/). All effects were estimated with random-effects models and the Restricted Maximum Likelihood (REML) estimator using the R-package metafor [@10.18637/jss.v036.i03], though with a variety of outcome variables: correlations (r), standardized mean differences (SMD), raw mean differences (MD), and risk differences (RD). ML1 transformed effects measured as odd ratios into standardized mean differences when meta-analyzing under the assumption that responses followed logistic distributions [@10.1037/1082-989X.8.4.448; @10.18637/jss.v036.i03]. Two projects (RRR5 and RRR7) used the Knapp and Hartung adjustment of the standard errors [@10.1002/sim.1482] and ML3 correlations were corrected for bias [@10.1037//0021-9010.74.3.469; @10.18637/jss.v036.i03]. After estimating effects, ML3 transformed correlations into eta-squared for reporting, which we did not. For each effect we estimated _I^2^_ and its 95% confidence interval. Confidence intervals were calculated in metafor which uses the Q-profile method [@10.1186/1471-2288-14-103].

In order to facilitate interpretation of our results, we estimated type I error and power of the Q-test of heterogeneity for each of the 37 effects under zero/small/medium/large heterogeneity (_I^2^_ = 0/25/50/75%  respectively). In addition, we approximated the probability density function of _I^2^_ across effects at each of these four heterogeneity levels and compared them with the observed frequency distribution of the observed _I^2^_ estimates of the 37 effects. Hence, five distributions of _I^2^_ were obtained; four simulated and one observed. To do so we simulated results of _I^2^_ for each effect given its number of studies (K), sample sizes of those studies (vector $N_k$), and each of the four heterogeneity levels (osf.io/gbf4u/). We directly simulated the distribution of _I^2^_ for correlation, standardized mean difference, and mean difference effect size measures, but not for risk ratios. We treated risk ratios as mean differences using the study sample sizes to compute study precision, because treating them as risk ratios would require strong assumptions on the probability of success in both treatment groups, assumptions which would greatly affect the outcomes of the simulation. For the same reason we treated the four effects of ML1 which were measured as odds ratios (and then transformed into standardized mean differences) as standardized mean differences.

As our concern was heterogeneity, for convenience we set the true effect size to zero in our simulations of heterogeneity. This should not affect the results for correlations or mean differences, as estimates of effect size and heterogeneity for these measures are statistically independent. For standardized mean differences we expect negligible effects on the results, because while these estimates of effect size are positively correlated to their standard errors, the within study variance $\sigma^2$ was kept constant across studies. As a sensitivity analysis we also ran all analyses assuming 'medium' effect sizes [@Cohen1988] and found the same average power at the different heterogeneity levels, see Appendix A (osf.io/bsjhu/).

In case the observed effect was a correlation, one run of a simulation proceeded as follows. First, we randomly sampled K Fisher-transformed true correlations $\rho_k$ from a normal distribution with mean 0 and heterogeneity (variance) $\tau^2$. Fisher's z-transformation is a normalizing correlation transformation that ranges from negative infinity to positive infinity, though except for extreme correlations it stays within the -1 to 1 range [@10.2307/2331838; @Fisher1921]. Second, for each of the K true Fisher-transformed correlations we sampled one Fisher-transformed observed correlation from a normal distribution with mean $\rho_k$ and variance $1 /(N_k - 3)$. Finally, we fitted a random-effects meta-analysis with REML and estimated _I^2^_ for that run. In the simulations, we varied the between-studies standard deviation $\tau$ between 0.000 and 0.50 in increments of 0.005, and used 1,000 runs at each step to approximate the distribution of _I^2^_ at that value for true heterogeneity.

For mean differences (and hence also for risk ratios) we assumed a within-study variance of one for both treatment and control groups, $\sigma^2_c = \sigma^2_t = 1$. For each run we then set the population mean of the control condition to 0 and sampled K treatment population means $\mu_k$ from $N(0, \tau)$. Subsequently, K sample means for both control and treatment conditions were sampled, with $\bar{x}_c \sim N(0, 1 / \sqrt{n_c})$ and $\bar{x}_t \sim N(\mu_k, 1 / \sqrt{n_t})$, where $n_t$ and $n_c$ were the observed treatment and control sample sizes for each study. Group variances were sampled using $s^2_c \sim \chi^2 (n_c - 1) / (n_c - 1)$ and  $s^2_t \sim \chi^2 (n_t - 1) / (n_t - 1)$. Finally, we fitted a random-effects meta-analysis with REML and estimated _I^2^_ for that run. For standardized mean differences (and odds ratios) we proceeded identically, except that in the final step we asked metafor to transform the effect size into a standardized mean difference in fitting the random-effects model. As with correlations, the distribution of _I^2^_ was approximated for values of $\tau$ from 0 to .5 in steps of .005, using 1,000 runs at each step.
